---
title: "Week 2 In-Class Exercise: Vectors & Subsetting"
subtitle: "GPH-GU 2182 · Statistical Programming in R"
author: "YOUR NAME & PARTNER NAME"
date: "September 9, 2026"
format:
  html:
    toc: true
    number-sections: true
    code-tools:
      source: true
      toggle: false
      caption: "Source Code"
editor: 
  markdown: 
    wrap: 72
---

::: callout-note
### Instructions

Work in pairs during class.

- 📥 [**Download
  `week-02-inclass.qmd`**](https://gph-2182.github.io/weeks/inclass/week-02-inclass.qmd)
  to open and edit in RStudio (or use the **`</> Source Code`** button
  at the top right).
- Render this Quarto document to HTML and push it to your GitHub
  repository by the end of class for participation credit.
:::

# Cycle 1: Vector Types & Implicit Coercion

## Type Predictions and Verification

In the code chunk below, write R code to check the `typeof()` of each of
the following expressions:

- `vec_a <- c(4, 7, FALSE)`
- `vec_b <- c(3L, 5.5, "control")`
- `vec_c <- c(TRUE, FALSE) + 5`

```{r type-checks}
# Create the vectors and check their types:
vec_a <- c(4, 7, FALSE)
vec_b <- c(3L, 5.5, "control")
vec_c <- c(TRUE, FALSE) + 5
typeof(vec_a)
typeof (vec_b)
typeof(vec_c)


```

## Public Health Prevalence Calculation

A study recorded the HIV viral load copies/mL for 8 participants:
`1200, 450, 0, 8900, 320, 0, 1500, 60`.

1.  Create a numeric vector named `viral_load` with these measurements.
2.  Use a logical comparison and the `mean()` function to calculate the
    **proportion of patients with undetectable viral load**
    ($\text{viral load} = 0$).

```{r viral-load}
# Calculate undetectable proportion:
viral_load <- c(1200, 450, 0, 8900, 320, 0, 1500, 60)
mean(viral_load==0)

```

------------------------------------------------------------------------

# Cycle 2: Creating Patterned Vectors

## Dosage Schedules with `seq()`

An intravenous antibiotic is administered every **8 hours** across a
**72-hour** hospital observation period.

Create a sequence named `dose_hours` starting at hour 0 up to hour 72
with increments of 8.

```{r dose-seq}
# Create dosage hours:
dose_hour <- seq (from =0, to = 72, by =8)

```

## Cohort Generation with `rep()`

A vaccine trial requires creating treatment labels for 12
participants: - 4 "Control" - 4 "Low-Dose" - 4 "High-Dose"

Generate this character vector using the `rep()` function and assign it
to `treatment_groups`.

```{r treatment-rep}
# Generate treatment labels:
treament_groups <- rep (c("Control", "Low Dose", "HighDose"), times = 4)

```

## Extracting Multiples with Modulo `%%`

Given the sequence `numbers <- 1:50`, extract all numbers that are
divisible by 7.

```{r modulo-filter}
numbers <- 1:50
# Extract multiples of 7:
numbers[numbers %%7 ==0]

```

------------------------------------------------------------------------

# Cycle 3: Subsetting & Mutating Vectors

## Glucose Filtering and Cleaning

A metabolic health study collected fasting blood glucose levels (mg/dL)
from 8 patients: `glucose <- c(95, 142, 180, 88, 210, 115, -1, 104)`

1.  Extract all values in the **pre-diabetes range**
    ($100 \le \text{glucose} \le 125$).
2.  The measurement `-1` is a missing value code. Replace all negative
    values in `glucose` with `NA`.
3.  Extract all valid readings ($\ge 126$, excluding `NA`) indicating
    **diabetes**.

```{r glucose-subset}
glucose <- c(95, 142, 180, 88, 210, 115, -1, 104)

# 1. Pre-diabetes subset:
glucose[100 <= glucose & glucose <=125]
# 2. In-place replacement of negative values with NA:
glucose [glucose == -1] <- NA

# 3. Diabetes subset (>= 126, non-NA):
glucose [glucose >=126& !is.na(glucose)]

```

------------------------------------------------------------------------

# Cycle 4: Sorting, Ranking & AI Verification

## Patient Rankings

A clinic recorded wait times (minutes) for 5 emergency room patients:
`wait_times <- c(45, 12, 85, 30, 45)`

1.  Sort the wait times in **descending order** (longest to shortest).
2.  Compute the ranks of each patient's wait time using
    `rank(..., ties.method = "first")`.

```{r wait-times}
wait_times <- c(45, 12, 85, 30, 45)

# 1. Descending sort:
sort (wait_times, decreasing = TRUE)
# 2. First-appearance ranking:
rank (wait_times, ties.method = "first")
```

## AI Corner: Spotting Vector Recycling Errors

An AI assistant was asked to calculate the Body Mass Index for 4
patients with heights (in meters) `c(1.70, 1.85, 1.62, 1.78)` and
weights (in kg) `c(65, 82)`.

The AI produced this code:

```{r ai-recycling}
# AI Generated code:
heights <- c(1.70, 1.85, 1.62, 1.78)
weights <- c(65, 82)
bmi_calc <- weights / (heights^2)
```

### Questions:

1.  What did R do when computing `bmi_calc`? Why is the result
    clinically invalid?
2.  How would you write a defensive check in R to prevent vector
    recycling when calculating BMI?
    
*(Write your explanation and defensive check below)*


When computing bmi_calc as written above, R will perform element-wise division. That is, the person with height = 1.70 will have a BMI with the weight 65; the person with height = 1.85 will have a BMI with weight = 82; the person with height = 1.62 will have a BMI with weight = 65; the person with height = 1.78 will have a BMI with the weight = 65). In short, given that the weights vector is length 2 and heights vector is length 5, R will 'recycle' the two values in weights so that the 2 elements of heights still have a corresponding weight for the calculation. 

This is clinically invalid unless the person with height 1.62 actually has a weight = 65 and the person with the height 1.78 actually has a weight =82. Otherwise, the recycling method that R will use will assign incorrect weights for 2 people, which render incorrect BMI values. 

A defensive check would be to first check if the length of the height vector is the same as the length of the weight vector:
```{r}
 if (!length(heights) == length(weights)) print ("not enough info")
#this should print "not enough info" if the height and length vectors are NOT equivalent length. 

```



------------------------------------------------------------------------

## Submission Checklist

- [ ] Both your name and your partner's name are in the `author` field
  at the top.
- [ ] Click **Render** in RStudio and verify the HTML compiles cleanly.
- [ ] Commit and push your changes to your GitHub repository.
